---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    pandoc_args: [
    "+RTS", "-K64m",
    "-RTS"
    ]
---


## Loading and preprocessing the data
```{r}
  if (!file.exists("activity.csv")) {
    if (!file.exists("activity.zip")) {
      print("ERROR: Cannot open file. Please check that you're running in the repo directory")
    } else {
      unzip("activity.zip")
    }
  }
  activity_data_raw <- read.csv("activity.csv")
```



## What is mean total number of steps taken per day?
####Calculate the total number of steps taken each day.  

1. Load the dplyr library to make sure it's available

2. Use dplyr functions group_by & summarize to sum steps for each day 
```{r}
library(dplyr)
byday <- group_by(activity_data_raw, date)
steps_byday <- summarize(byday, step_sum = sum(steps, na.rm=TRUE))
```

####Make a histogram of the total number of steps taken each day
```{r}
hist(steps_byday$step_sum, main="Histogram of sum of steps by day", xlab="Sum of Daily Steps")
```

####Calculate & report the mean number of steps per day
```{r}
mean(steps_byday$step_sum)
```
####Calculate & report the median number of steps per day
```{r}
median(steps_byday$step_sum)
```

##What is the average daily activity pattern?

####Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

First, group by interval & find the average number of steps for each interval across days
```{r}
byinterval <- group_by(activity_data_raw, interval)
steps_byinterval <- summarize(byinterval, step_avg = mean(steps, na.rm=TRUE))
```
Then, make the time series plot of inverval (x-axis) vs average steps (y-axis)
```{r}
plot(steps_byinterval$interval, steps_byinterval$step_avg, type="l", 
	xlab="Time Interval", ylab="Steps averaged across all days")
```

####Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

First, reorder the data frame so it's in order from most to least average steps
```{r}
steps_ordered <- arrange(steps_byinterval, desc(step_avg))
```

Then, print the interval value of the top entry in the re-ordered data frame.  That will be the interval with the highest average steps.
```{r}
head(select(steps_ordered, interval), 1)
```

##Imputing missing values

####Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
Calling summary on an object will let you know the number of NA values in the object
```{r}
summary(activity_data_raw)

```

####Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Any attempt to fill in missing data always has major downsides - even if you're using related values found in the dataset, you're still making major assumptions that may very well be false.  However, in cases where you must do this, you should always look for patterns in what data is missing as well as look at what analysis you're going to do on the modified data so you can best mitigate impact of potentially "bad" data.

In this case, any day with an NA value for any interval actually has NA for every interval that day.  So, we're not just missing periodic elements within a day, but we missing the entire day's data in each case.  Given that fact, I opted to replace each NA with the average value for that 5-minute interval since, if you were to use the mean for that day, all of those days (8 in total) would be 0 for every interval.  That seems like a less realistic replacement method than using the average for that interval from all other days measured since  using the daily mean, 0, would make the data show those people were completely still for 8 days of the dataset.

####Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_data_complete <- activity_data_raw

for (i in 1:length(activity_data_complete$steps)) {
	#Logic here is if the current steps value is NA, find the average steps value (grouped by interval) 
	#for the same interval as the current iteration & replace the NA value with the average steps for this interval
	if (is.na(activity_data_complete$steps[i])) {
		  activity_data_complete$steps[i] <- filter(steps_byinterval, 
		  	interval==activity_data_complete$interval[i]) %>% select (step_avg)
	}	
}
#This logic turns step into a list
#Flatten the list so summarize can be used in later steps
activity_data_complete$steps <- unlist(activity_data_complete$steps)
```
####Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 
Make a histogram

```{r}
byday_c <- group_by(activity_data_complete, date)
steps_byday_c <- summarize(byday_c, step_sum = sum(steps))
hist(steps_byday_c$step_sum, main="Histogram of sum of steps by day", xlab="Sum of Daily Steps")
```

Calculate mean
```{r}
mean(steps_byday_c$step_sum)
```

Calculate median
```{r}
median(steps_byday_c$step_sum)
```


####Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Because I chose to replace NAs with the average step value for a given interval, it did change the mean/median of the resulting data set.  (The original computation had used 0 for all of those NA values; so, if I had instead replaced the NAs with the mean of each day (0, in all cases) these values would have stayed the same.)  However, by replacing misisng values with the mean for the interval, it changed the mean & median values; they now match.

## Are there differences in activity patterns between weekdays and weekends?
####Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
#Add a column with day of week
activity_data_complete <- mutate(activity_data_complete, day=weekdays(as.Date(activity_data_complete$date)))
#Convert day of week to weekday or weekend
for (i in 1:length(activity_data_complete$day)) {
	#Logic here is if the current day value is Saturday or Sunday, replace with weekend
	#Otherwise, replace with weekday
	if ((activity_data_complete$day[i] == "Sunday") | (activity_data_complete$day[i] == "Saturday")) {
		  activity_data_complete$day[i] <- "weekend"
	}	else {
		activity_data_complete$day[i] <- "weekday"
	}
}
#Make it a factor
activity_data_complete$day <- as.factor(activity_data_complete$day)
```
####Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


```{r}
#First, group by day/interval, then average the number of steps for each grouping
step_byday_byinterval <- activity_data_complete %>% group_by(day, interval) %>% 
	summarize(step_avg = mean(steps))
```
Then, make the time series plot of inverval (x-axis) vs average steps (y-axis)
```{r}
library(ggplot2)
g <- ggplot(step_byday_byinterval, aes(interval, step_avg)) + geom_line() + facet_grid(day~.) 
g <- g + labs(y = "Number of Steps")
g <- g + labs(x = "Interval")
print(g)
```